Indicators for Ecosystem Condition in Norway

### 2023-09-29 NDVI_seminat.Rmd
cut short friday

# NDVI (greenness) indicator, semi-natural ecosystems  {#NDVI-indicator-seminat}

_Author and date:_
Joachim Töpper, Tessa Bargmann

@@

<!-- Load all you dependencies here -->
```{r setup, include=FALSE}
library(downloader)
library(data.table)
library(lubridate)
library(sf)
library(plyr)
library(stringr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra")
library(ggridges)
library(tmap)
library(tmaptools)
library(betareg)
library(StepBeta)
library(glmmTMB)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

<!-- Don't remove these three html lines -->
<br />
<br />
<hr />

## Introduction
The Normalized difference vegetation index (NDVI) can be used to describe the greenness of an area, and may thus represent a proxy for the amount of chlorophyll and plant productivity in the given area. Different ecosystem types naturally have greenness signals of different intensity but also with a certain variance. Deviations beyond the natural variation in either direction may indicate a reduction in ecological condition of a given ecosystem. E.g. lower than normal NDVI may indicate excessive disturbance, while higher than normal NDVI may indicate regrowth and transitions towards more densely vegetated systems. In recent assessments of ecological condition for forest and mountain ecosystems, two different NDVI-based indicators were developed and applied. For forests, NDVI-values across the country were compared to a spatial reference NDVI-signal from protected areas (Framstad et al. 2020). For mountains, the extent and consistence of the temporal change of NDVI-values over the last twenty years was assessed (Framstad et al 2021). In both cases, the NDVI-product from the MODIS satellite reflectance imagery was used based on available maps of spatial ecosystem extent in Norway (i.e. national maps for the forest and mountain ecosystems).

Here, we explore the potential for developing an NDVI-indicator for ecological condition representing open semi-natural ecosystems. For this task, two initial challenges can be identified early on:
1) We don't yet have a national map for semi-natural ecosystems, i.e. we don't know the exact locations of all semi-natural areas in Norway, which is a prerequisite to extract satellite data for the right locations and ecosystems, and thus achieve a good and unbiased spatial representation for the assessment area.
2) The spatial extent of the single semi-natural ecosystem occurrences (hereafter 'polygons') includes many areas which are way smaller than the pixel size in the MODIS imagery (250 x 250 m) which was used for forests and mountains and has a time series back to year 2000.

In this draft development for a NDVI indicator for ecological condition in semi-natural  systems we approach these challenges in the following ways:
1) We use ecosystem occurrence data from NiN-mapping to guide the extraction of satellite NDVI data for pixels representing the right ecosystem. However, the ecosystem occurrences in the NiN data are highly spatially biased and thus cannot provide a good and unbiased spatial representation for the assessment area. Our work therefore represents a concept and method development which may be applied in an assessment of ecological condition once spatially unbiased occurrence data or a national map of semi-natural ecosystems are available.
2) We include Sentinel-2 imagery for semi-natural ecosystems as the resolution for the bands underlying the NDVI product here is 10 x 10 m which is smaller than almost all semi-natural polygons in the NiN-data. However, since Sentinel-2 only goes back to 2015, we lose much of the time series aspect as compared to the MODIS data. In addition, we include Landsat imagery which goes back to 1984 and has a resolution of 30 x 30 m, but also has more variable data due to less frequent seasonal measurements. We also still explore NDVI data from MODIS (see more details below).

## About the underlying data
In the NDVI project for semi-natural systems, we use two kinds of data (coming as four datasets) for building indicators for ecological condition:

@@

(1) NiN ecosystem data:
The NiN data in this study contain data about the ecosystem categories (cf. Halvorsen et al. 2020) present at mapped sites as wall as their spatial delimitation (in the form of polygons). Since we use NiN data that were collected following 'Miljødirektoratets instruks', these data also include information on the ecological condition and underlying condition variabels of the mapped sites (see the Miljødirektoratets mapping manual for details).
Polygon sizes in these NiN data may vary from <100 sqm to the dimension of sqkm's.

@@

### Representativity in time and space
For semi-natural systems, the NiN data contain 24494 semi-natural polygons of known area size across mainland Norway. However, NiN-mapping is not planned and performed in a spatially representative way and thus both the ecosystem occurrences and their spatial extent have to be treated as spatially biased. The 24494 semi-natural polygons are distributed across regions and major ecosystem types in the following way:

                  T31   T32   T33   T34   T40   T41
Northern Norway   1194  2081  14    1     15    0
Central Norway    641   2518  210   2338  97    0
Western Norway    132   2922  94    2200  66    54
Eastern Norway    1064  4488  10    10    371   54
Southern Norway   234   2317  112   749   111   140

In addition, there is 10 T31s, 78 T32s, 15 T33s, 59 T34s, and 1 T40 that are not assigned to any region.

@@

### Temporal coverage
The currently available NiN data on semi-natural ecosystems span a time period from 2018-2021 and thus represent a contemporary sample of ecosystem occurrences. Note that none of the polygons in the NiN-data has been visited more than once during that period.

Available Sentinel-2 satellite NDVI data span the years 2015-2022.
Available MODIS satellite NDVI data span the years 2000-2022.
Available Landsat satellite NDVI data span the years 1984-2021.

## Collinearities with other indicators
The NDVI signal may likely be co-linear with the re-growth indicator based on LiDAr derived canopy height in abandoned semi-natural areas under shrub and tree encroachment.

## Reference state and values
### Reference state and scaling values
This indicator project is mainly exploratory, mapping the possibilities of and lacking prerequisites for satellite derived NDVI data to inform ecological condition assessments. Therefore, we lack a concise strategy for how to define the reference state, especially prior to having access to an ecosystem map for semi-natural ecosystems (cf. Jakobsson et al. 2020, Töpper & Jakobsson 2021). In this exploratory work, we will use the condition classification in the NiN-data to explore (1) how NDVI connects to semi-natural ecology and condition and (2) its applicability for defining a reference state. We will use statistical models to test whether or not NDVI varies systematically with condition and across time.

## Uncertainties
Given a condition index can be achieved, we can calculate a mean indicator value (after scaling) for every region (or any other delimited area of interest) as well as its corresponding standard error and standard deviation as a measure of spatial uncertainty for a geographical area (see Töpper & Jakobsson 2021).

@@

Halvorsen, R., Skarpaas, O., Bryn, A., Bratli, H., Erikstad, L., Simensen, T., & Lieungh, E. (2020). Towards a systematics of ecodiversity: The EcoSyst framework. Global Ecology and Biogeography, 29(11), 1887-1906. doi:10.1111/geb.13164

@@

Miljødiretoratets mapping manual: https://www.miljodirektoratet.no/publikasjoner/2022/januar/kartleggingsinstruks-kartlegging-av-terrestriske-naturtyper-etter-nin/

@@

## Analyses
### Data sets
The analyses in this document make use of the following data sets:
-NiN data (Naturtyper etter Miljødirektoratets instruks)
-Sentinel-2 NDVI data
-MODIS NDVI data
-Landsat NDVI data

#### NiN data and geometry for Norway and the five regions
We read the NiN data for this document from a cache at the GitHub repository...
```{r}
# Add NiN data from cache
nin <- readRDS(paste0(here::here(), "/data/cache/nin.RDS"))
```

...but it can also be directly downloaded from Miljødirektoratets kartkatalog (https://kartkatalog.miljodirektoratet.no/Dataset)
```{r, echo = T, eval = F}
url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
download(url, dest = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode = "wb")
unzip("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip",
  exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data"
)
st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
  layer = "naturtyper_nin_omr"
)
```
And we upload maps for Norway and the five regions: Southern, Western, Eastern, Central, and Northern Norway
```{r, message=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

@@

#### NDVI data
All NDVI data are calculated mean NDVI values for each respective NiN polygon from each available Sentinel-2, MODIS, or Landsat image in Google Earth Engine (GEE). 
```{r, echo = F}
# downloading the unprocessed NDVI data from a data-cache et the NINA server
df.s <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
df.m <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
df.l <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
```
##### Sentinel-2 NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4
The image collection contains Sentinel imagery from June, July and August 2015-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r, echo = T, eval = F}
# Import Sentinel NDVI Data
df.s <- list.files("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names = TRUE) %>%
  map_df(~ fread(.))
```

##### MODIS NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84
The image collection contains MODIS imagery from June, July and August 2000-2022

@@

##### Landsat NDVI data
The GEE code can be seen here: https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7
The image collection contains Landsat imagery from June, July and August 1984-2022
To not exceed GEE memory limits, the exported files had to be iterated over a grid which resulted in 42 separate csv files. This script merges them and then merges the dataframe to the NiN data.
```{r, echo = T, eval = F}
# Import Landsat NDVI Data
# Set up conditional file paths
dir <- substr(getwd(), 1, 2)

@@

### Data handling
#### NiN data
```{r, echo = T, eval = F}
# fixing variable- and ecosystem-names with special characters
colnames(nin)
colnames(nin)[c(3, 8, 17, 26, 31, 33, 34)] <- c(
  "hovedoekosystem", "kartleggingsaar", "noyaktighet",
  "omraadenavn", "uk_naertruet", "uk_sentraloekosystemfunksjon",
  "uk_spesieltdaarligkartlagt"
)
unique(nin$hovedoekosystem)

@@

```{r, echo = F}
# load processed NiN-data from cache
nin.seminat <- readRDS(paste0(here::here(), "/data/cache/nin.seminat.RDS"))
nin.seminat2 <- readRDS(paste0(here::here(), "/data/cache/nin.seminat2.RDS"))
```

#### NDVI data
```{r, results='hide', message=FALSE}
## Sentinel-2
# join nin.seminat & Sentinel NDVI data
SentinelNDVI.seminat <- full_join(nin.seminat, df.s, by = "id")
# summary(SentinelNDVI.seminat)
SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(SentinelNDVI.seminat)
  )
# summary(SentinelNDVI.seminat)
# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(!is.na(hovedtype))
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(!is.na(mean))
# summary(SentinelNDVI.seminat)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.seminat)
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.seminat)
# split date into year, month & day
SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# add column for sub-ecosystem types
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% mutate(
  subtype = substring(ninkartleggingsenheter, 4),
  subtype = str_remove(subtype, "-"),
  subtype = str_remove(subtype, "-")
)
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

@@

@@

# Where are the seminat sites?
```{r}
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.seminat) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T31 - boreal heath",
      "T32 - semi-natural grasslands",
      "T33 - semi-natural coastal grasslands",
      "T34 - coastal heathlands",
      "T40/41 - previously heavily altered grassland"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Semi-natural ecosystems (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

# How are polygon sizes distributed?
```{r}
summary(SentinelNDVI.seminat$area_meters)
hist(SentinelNDVI.seminat$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```
semi-natural polygons stretch a large number of sizes from just above 100 sqm to almost 3 million sqm. 75% of all polygons are below 15297 sqm.

# How does NDVI vary over the years
```{r}
SentinelNDVI.seminat %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```
2022 stands out with the highest NDVI values missing.
WE omit 2022 for this analysis
```{r}
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(year != "2022")
```

# How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)
```{r warning=FALSE}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")

>>>>

---
editor_options: 
  markdown: 
    wrap: 72
---

# NDVI- Semi-natural Ecosystems {#NDVI-indicator-seminat}

<br /> Norwegian name: **Primærproduksjon** <br />

*Author and date:*

Joachim Töpper, Tessa Bargmann

@@

<!-- Load all you dependencies here -->

```{r setup, include=FALSE}
library(downloader)
library(data.table)
library(lubridate)
library(sf)
library(plyr)
library(stringr)
library(tidyverse)
library(RColorBrewer)
library("gridExtra")
library(ggridges)
library(tmap)
library(tmaptools)
library(betareg)
library(StepBeta)
library(glmmTMB)
library(knitr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

<!-- Don't remove these three html lines -->

<br /> <br />

<hr />

## Introduction {#intro-ndvis}

The Normalized difference vegetation index (NDVI) can be used to
describe the greenness of an area, and may thus represent a proxy for
the amount of chlorophyll and plant productivity in the given area.
Different ecosystem types naturally have greenness signals of different
intensity but also with a certain variance. Deviations beyond the
natural variation in either direction may indicate a reduction in
ecological condition of a given ecosystem. E.g. lower than normal NDVI
may indicate excessive disturbance, while higher than normal NDVI may
indicate regrowth and transitions towards more densely vegetated
systems. In recent assessments of ecological condition for forest and
mountain ecosystems, two different NDVI-based indicators were developed
and applied. For forests, NDVI-values across the country were compared
to a spatial reference NDVI-signal from protected areas (Framstad et al.
2020). For mountains, the extent and consistence of the temporal change
of NDVI-values over the last twenty years was assessed (Framstad et al
2021). In both cases, the NDVI-product from the MODIS satellite
reflectance imagery was used based on available maps of spatial
ecosystem extent in Norway (i.e. national maps for the forest and
mountain ecosystems).

Here, we explore the potential for developing an NDVI-indicator for
ecological condition representing open semi-natural ecosystems. For this
task, two initial challenges can be identified early on:

1.  We don't yet have a national map for semi-natural ecosystems, i.e.
    we don't know the exact locations of all semi-natural areas in
    Norway, which is a prerequisite to extract satellite data for the
    right locations and ecosystems, and thus achieve a good and unbiased
    spatial representation for the assessment area.
2.  The spatial extent of the single semi-natural ecosystem occurrences
    (hereafter 'polygons') includes many areas which are way smaller
    than the pixel size in the MODIS imagery (250 x 250 m) which was
    used for forests and mountains and has a time series back to year
    2000.

In this draft development for a NDVI indicator for ecological condition
in semi-natural systems we approach these challenges in the following
ways:

1.  We use ecosystem occurrence data from NiN-mapping to guide the
    extraction of satellite NDVI data for pixels representing the right
    ecosystem. However, the ecosystem occurrences in the NiN data are
    highly spatially biased and thus cannot provide a good and unbiased
    spatial representation for the assessment area. Our work therefore
    represents a concept and method development which may be applied in
    an assessment of ecological condition once spatially unbiased
    occurrence data or a national map of semi-natural ecosystems are
    available.
2.  We include Sentinel-2 imagery for semi-natural ecosystems as the
    resolution for the bands underlying the NDVI product here is 10 x 10
    m which is smaller than almost all semi-natural polygons in the
    NiN-data. However, since Sentinel-2 only goes back to 2015, we lose
    much of the time series aspect as compared to the MODIS data. In
    addition, we include Landsat imagery which goes back to 1984 and has
    a resolution of 30 x 30 m, but also has more variable data due to
    less frequent seasonal measurements. We also still explore NDVI data
    from MODIS (see more details below).

## About the underlying data {#underlying-data-ndvis}

In the NDVI project for semi-natural systems, we use two kinds of data
(coming as four datasets) for building indicators for ecological
condition:

@@

(1) NiN ecosystem data: The NiN data in this study contain data about
    the ecosystem categories (cf. Halvorsen et al. 2020) present at
    mapped sites as wall as their spatial delimitation (in the form of
    polygons). Since we use NiN data that were collected following
    'Miljødirektoratets instruks', these data also include information
    on the ecological condition and underlying condition variabels of
    the mapped sites (see the Miljødirektoratets mapping manual for
    details). Polygon sizes in these NiN data may vary from \<100 sqm to
    the dimension of sqkm's.

@@

### Representativity in time and space {#representativity-ndvis}

For semi-natural systems, the NiN data contain 24494 semi-natural
polygons of known area size across mainland Norway. However, NiN-mapping
is not planned and performed in a spatially representative way and thus
both the ecosystem occurrences and their spatial extent have to be
treated as spatially biased. The 24494 semi-natural polygons are
distributed across regions and major ecosystem types in the following
way:

| Region          | T31  | T32  | T33 | T34  | T40 |
|-----------------|------|------|-----|------|-----|
| Northern Norway | 1194 | 2081 | 14  | 1    | 15  |
| Central Norway  | 641  | 2518 | 210 | 2338 | 97  |
| Western Norway  | 132  | 2922 | 94  | 2200 | 66  |
| Eastern Norway  | 1064 | 4488 | 10  | 10   | 371 |
| Southern Norway | 234  | 2317 | 112 | 749  | 111 |

In addition, there are 10 T31s, 78 T32s, 15 T33s, 59 T34s, and 1 T40
that are not assigned to any region.

@@

### Temporal coverage {#temp-cov-ndvis}

The currently available NiN data on semi-natural ecosystems span a time
period from 2018-2021 and thus represent a contemporary sample of
ecosystem occurrences. Note that none of the polygons in the NiN-data
has been visited more than once during that period.

Available Sentinel-2 satellite NDVI data span the years 2015-2022.

Available MODIS satellite NDVI data span the years 2000-2022.

Available Landsat satellite NDVI data span the years 1984-2021.

## Collinearities with other indicators {#coll-ndvis}

The NDVI signal may likely be co-linear with [the re-growth indicator](https://ninanor.github.io/ecosystemCondition/gjengroing.html)
based on LiDAR derived canopy height in abandoned semi-natural areas
under shrub and tree encroachment.

## Reference state and values {#ref-state-ndvis}

### Reference state and scaling values {#ref-state2-ndvis}

This indicator project is mainly exploratory, mapping the possibilities
of and lacking prerequisites for satellite derived NDVI data to inform
ecological condition assessments. Therefore, we lack a concise strategy
for how to define the reference state, especially prior to having access
to an ecosystem map for semi-natural ecosystems (cf. Jakobsson et al.
2020, Töpper & Jakobsson 2021). In this exploratory work, we will use
the condition classification in the NiN-data to explore (1) how NDVI
connects to semi-natural ecology and condition and (2) its applicability
for defining a reference state. We will use statistical models to test
whether or not NDVI varies systematically with condition and across
time.

## Uncertainties {#uncertainties-ndvis}

Given a condition index can be achieved, we can calculate a mean
indicator value (after scaling) for every region (or any other delimited
area of interest) as well as its corresponding standard error and
standard deviation as a measure of spatial uncertainty for a
geographical area (see Töpper & Jakobsson 2021).

@@

Halvorsen, R., Skarpaas, O., Bryn, A., Bratli, H., Erikstad, L.,
Simensen, T., & Lieungh, E. (2020). Towards a systematics of
ecodiversity: The EcoSyst framework. Global Ecology and Biogeography,
29(11), 1887-1906. <doi:10.1111/geb.13164>

@@

Miljødiretoratets mapping manual:
<https://www.miljodirektoratet.no/publikasjoner/2022/januar/kartleggingsinstruks-kartlegging-av-terrestriske-naturtyper-etter-nin/>

@@

## Analyses {#analyses-ndvis}

### Data sets {#datasets-ndvis}

The analyses in this document make use of the following data sets:

- NiN data (Naturtyper etter Miljødirektoratets instruks) 
- Sentinel-2 NDVI data 
- MODIS NDVI data 
- Landsat NDVI data

#### NiN data and geometry for Norway and the five regions

We read the NiN data for this document from a cache at the GitHub
repository...

```{r}
# Add NiN data from cache
nin <- readRDS(paste0(here::here(), "/data/cache/nin.RDS"))
```

...but it can also be directly downloaded from Miljødirektoratets
kartkatalog (<https://kartkatalog.miljodirektoratet.no/Dataset>)

```{r, echo = T, eval = F}
url <- "https://nedlasting.miljodirektoratet.no/Miljodata//Naturtyper_nin/FILEGDB/4326/Naturtyper_nin_0000_norge_4326_FILEGDB.zip"
download(url, dest = "P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip", mode = "wb")
unzip("P:/41201785_okologisk_tilstand_2022_2023/data/Naturtyper_nin_0000_norge_4326_FILEGDB.zip",
  exdir = "P:/41201785_okologisk_tilstand_2022_2023/data/nin_data"
)
st_layers("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb")
nin <- st_read("P:/41201785_okologisk_tilstand_2022_2023/data/nin_data/Naturtyper_nin_0000_norge_4326_FILEGDB.gdb",
  layer = "naturtyper_nin_omr"
)
```

And we upload maps for Norway and the five regions: Southern, Western,
Eastern, Central, and Northern Norway

```{r, message=FALSE}
# Import region- og Norgeskart, make sure they have the same geometry as the nin data
nor <- st_read("data/outlineOfNorway_EPSG25833.shp") %>%
  st_as_sf() %>%
  st_transform(crs = st_crs(nin))

@@

#### NDVI data

All NDVI data are calculated mean NDVI values for each respective NiN
polygon from each available Sentinel-2, MODIS, or Landsat image in
Google Earth Engine (GEE).

<!-- RUNTIME 30 sec -->
```{r, echo = F}
# downloading the unprocessed NDVI data from a data-cache et the NINA server
df.s <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Sentinel.raw.RDS")
df.m <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Modis.raw.RDS")
df.l <- readRDS("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/ndvi.Landsat.raw.RDS")
```

##### Sentinel-2 NDVI data

The GEE code can be seen here:
<https://code.earthengine.google.com/2ceb0c3e03adade9e6f6d0903184b8c4>
The image collection contains Sentinel imagery from June, July and
August 2015-2022 To not exceed GEE memory limits, the exported files had
to be iterated over a grid which resulted in 42 separate csv files. This
script merges them and then merges the dataframe to the NiN data.

```{r, echo = T, eval = F}
# Import Sentinel NDVI Data
df.s <- list.files("/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/NDVI_åpenlavland/NDVI_data_Sentinel/", pattern = "*.csv", full.names = TRUE) %>%
  map_df(~ fread(.))
```

##### MODIS NDVI data

The GEE code can be seen here:
<https://code.earthengine.google.com/efb84013701f1d5f6e1e81345f389b84>
The image collection contains MODIS imagery from June, July and August
2000-2022

@@

##### Landsat NDVI data

The GEE code can be seen here:
<https://code.earthengine.google.com/da8a9279238ef26d14be08a43788b6b7>
The image collection contains Landsat imagery from June, July and August
1984-2022 To not exceed GEE memory limits, the exported files had to be
iterated over a grid which resulted in 42 separate csv files. This
script merges them and then merges the dataframe to the NiN data.

```{r, echo = T, eval = F}
# Import Landsat NDVI Data
# Set up conditional file paths
dir <- substr(getwd(), 1, 2)

@@

### Data handling {#data-handling-ndvis}

#### NiN data {#nin-data-ndvis}

```{r, echo = T, eval = F}
# fixing variable- and ecosystem-names with special characters
colnames(nin)
colnames(nin)[c(3, 8, 17, 26, 31, 33, 34)] <- c(
  "hovedoekosystem", "kartleggingsaar", "noyaktighet",
  "omraadenavn", "uk_naertruet", "uk_sentraloekosystemfunksjon",
  "uk_spesieltdaarligkartlagt"
)
unique(nin$hovedoekosystem)

@@

<!-- RUNTIME 2 sec -->
```{r, echo = F}
# load processed NiN-data from cache
nin.seminat <- readRDS(paste0(here::here(), "/data/cache/nin.seminat.RDS"))
nin.seminat2 <- readRDS(paste0(here::here(), "/data/cache/nin.seminat2.RDS"))
```

#### NDVI data {#ndvi-data-ndvis}

<!-- RUNTIME 4 min. Caching it. -->
```{r, results='hide', message=FALSE, eval=F}
## Sentinel-2
# join nin.seminat & Sentinel NDVI data
SentinelNDVI.seminat <- full_join(nin.seminat, df.s, by = "id")
# summary(SentinelNDVI.seminat)
SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
  mutate(
    hovedoekosystem = as.factor(hovedoekosystem),
    hovedtype = as.factor(hovedtype),
    ninkartleggingsenheter = as.factor(ninkartleggingsenheter),
    lokalitetskvalitet = as.factor(lokalitetskvalitet),
    tilstand = as.factor(tilstand),
    area_meters = st_area(SentinelNDVI.seminat)
  )
# summary(SentinelNDVI.seminat)
# get rid of NAs (i.e. NDVI cells that were not in semi-natural polygons)
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(!is.na(hovedtype))
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(!is.na(mean))
# summary(SentinelNDVI.seminat)
# get rid of any nin-polygons smaller than the Sentinel grid cell size (100 sqm)
dim(SentinelNDVI.seminat)
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(as.numeric(area_meters) >= 100)
dim(SentinelNDVI.seminat)
# split date into year, month & day
SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
  dplyr::mutate(
    year = lubridate::year(date),
    month = lubridate::month(date),
    day = lubridate::day(date)
  )
# add column for sub-ecosystem types
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% mutate(
  subtype = substring(ninkartleggingsenheter, 4),
  subtype = str_remove(subtype, "-"),
  subtype = str_remove(subtype, "-")
)
# we are using max NDVI per year in every NiN polygon
SentinelNDVI.seminat <- SentinelNDVI.seminat %>%
  group_by(id, year) %>%
  filter(mean == max(mean, na.rm = TRUE))

@@

```{r cahing data ndvis, eval=FALSE, include=F}
saveRDS(LandsatNDVI.seminat, 
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/LandsatNDVI.seminat.rds")

saveRDS(ModisNDVI.seminat, 
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/ModisNDVI.seminat.rds")

saveRDS(SentinelNDVI.seminat, 
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/SentinelNDVI.seminat.rds")
```

```{r, echo=F}
LandsatNDVI.seminat <- readRDS(
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/LandsatNDVI.seminat.rds")

ModisNDVI.seminat <- readRDS( 
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/ModisNDVI.seminat.rds")

SentinelNDVI.seminat <- readRDS( 
        "/data/P-Prosjekter2/41201785_okologisk_tilstand_2022_2023/data/cache/SentinelNDVI.seminat.rds")
```

@@

**Where are the seminat sites?**

```{r where-are-seminat-sites, }
tm_shape(regnor) +
  tm_fill("GID_0", labels = "", title = "", legend.show = FALSE) +
  tm_borders() +
  tm_shape(SentinelNDVI.seminat) +
  tm_dots("hovedtype", midpoint = NA, palette = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE), scale = 2.5, legend.show = FALSE) +
  tm_add_legend(
    type = "symbol",
    col = tmaptools::get_brewer_pal("YlOrRd", 5, plot = FALSE),
    labels = c(
      "T31 - boreal heath",
      "T32 - semi-natural grasslands",
      "T33 - semi-natural coastal grasslands",
      "T34 - coastal heathlands",
      "T40/41 - previously heavily altered grassland"
    ),
    title = "Major ecosystem type",
    size = 3
  ) +
  tm_layout(
    main.title = "Semi-natural ecosystems (NiN, Miljødirektoratets instruks)",
    main.title.size = 1.2, legend.position = c("right", "bottom"),
    legend.text.size = 1.3, legend.title.size = 1.4
  )
```

**How are polygon sizes distributed?**

```{r}
summary(SentinelNDVI.seminat$area_meters)
hist(SentinelNDVI.seminat$area_meters, xlim = c(0, 20000), breaks = 50000)
abline(v = 100, lty = 2)
```

semi-natural polygons stretch a large number of sizes from just above
100 sqm to almost 3 million sqm. 75% of all polygons are below 15297
sqm.

**How does NDVI vary over the years**

```{r}
SentinelNDVI.seminat %>%
  ggplot(aes(x = year, y = mean)) +
  geom_point() +
  facet_grid(tilstand ~ hovedtype)
```

2022 stands out with the highest NDVI values missing. WE omit 2022 for
this analysis

```{r}
SentinelNDVI.seminat <- SentinelNDVI.seminat %>% filter(year != "2022")
```

**How do NDVI values vary between major and basic ecosystem types for sites in good condition? (using only NDVI data matching NiN-mapping years)**

```{r warning=FALSE}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = hovedtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey")

SentinelNDVI.seminat$subtype <- factor(SentinelNDVI.seminat$subtype, levels = c("C1", "C2", "C3", "C4", "C5", "C6", "C7", "C8", "C9", "C10", "C11", "C12", "C13", "C14", "C15", "C16", "C17", "C18", "C19", "C20", "C21", "E1", "E2", "E3", "E4", "E5", "E6", "E7"))
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  ggplot(aes(x = subtype, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2, size = 6))
```

NDVI varies between major ecosystem types with the highest values in
semi-natural grasslands (T32) and in previously intensively used and
altered grasslands that today resemble a semi-natural grassland (T40,
T41), lowest values in coastal heathlands (T34). Other ecosystem types:
Boreal heath (T31), semi-natural coastal grasslands (T33). The overlaps
between the ecosystem types are considerable. With respect to basic
ecosystem types within the major types, NDVI appears to cover very
similar ranges across the subtypes, and those with deviating violin
plots have a smaller sample size. There is no obvious trend related to
the structuring environmental gradients withing the major types, and
also here the overlap is large.

**How does NDVI vary across regions in semi-naturals (major types) in good condition? (using only NDVI data matching NiN-mapping years)**

```{r warning=FALSE}
SentinelNDVI.seminat %>%
  filter(tilstand == "God") %>%
  filter(year == kartleggingsaar) %>%
  filter(!is.na(region)) %>%
  ggplot(aes(x = region, y = mean)) +
  geom_violin() +
  geom_point(size = 0.7, shape = 16, color = "grey") +
  facet_wrap(~hovedtype) +
  theme(axis.text.x = element_text(angle = -45, vjust = 0.5, hjust = 0.2))
```
